Combining information from photometric and spectroscopic data:\\ Don't waste your time in spectrophotometric calibration!

### 2014-08-21 tex/ms.tex
remove old model equations from sec2, add a list of tests, comment out figures from the old model

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a trivial modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N \left[\frac{[d_n - \mu_n(\theta)]^2}{[\sigma_n^2 + s^2]} + \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements ({\color{blue} BDJ: IS THIS TRUE?} {\color{red} DWH,
Currently I'm not using a jitter term, but I can and will.  Update:
having difficulty with a jitter term}) in what follows.  In this
formulation, if $\theta$ are the parameters of interest, the jitter
variance $s^2$ is a nuisance noise-model parameter.  The Gaussian
Process will be a generalization of this modification.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - f(\alpha)\ast\mu(\theta)
\\
C_{nn'} &\equiv& [\sigma_n^2 + s^2]\,\delta_{nn'} + K_\alpha(\lambda_n - \lambda_{n'})
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
the ``$\ast$'' operator is the pixel-wise multiplication operator,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian process is going to
pick up smaller-scale residuals away from \emph{that}.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

XXX Alternative XXX.
Because the residual column vector $\Delta$ of equation
\ref{eq:spectroscopicLF} is additive with respect to the mean model
$\mu$, sharp spectral features in the mean model (such as absorption
lines) can be artificially
diluted or enhanced, affecting the inferred physical parameters. To mitigate this we construct a
model where the residual column vector is taken to be the residual
\emph{from the calibration vector $f(\alpha)$} and is multiplicative
with respect to the mean model $\mu$.  That is,
\begin{eqnarray}\label{eq:spectroscopicLFalt} \Delta &\equiv&
\frac{d}{\mu(\theta)} - f(\alpha) \\
\sigma_{n} & = & \sigma'_n / \mu(\theta)
\quad ,
\end{eqnarray}
where $d$, $\mu(\theta)$ and $f(\alpha)$ are as before, and
$\sigma'_{n}$ are now the observational uncertainties

XXX Alternative 2 XXX.
Because the residual column vector $\Delta$ of equation
\ref{eq:spectroscopicLF} is additive with respect to the mean model
$\mu$, sharp spectral features in the mean model (such as absorption
lines) can be artificially diluted or enhanced, affecting the inferred
physical parameters. To mitigate this we construct a noise model where the
residual column vector is taken to be additive in the log and is
therefore multiplicative with respect to the mean model $\mu$.  That
is,
\begin{eqnarray}\label{eq:spectroscopicLFalt2} 
\Delta &\equiv&
\ln d - \ln \mu(\theta) - f(\alpha) \\
\sigma_{n} & = & \sigma'_n / d
\quad ,
\end{eqnarray}
where $d$, $\mu(\theta)$ and $f(\alpha)$ are as before, and
$\sigma'_{n}$ are now the observational uncertainties. In this
formulation the the mean model $\mu$ will be effectively multiplied by
$e ^{f(\alpha)}$ ({\color{red} could we just add
this to the covariance matrix with a single long quadratic ?}).  This
logarithmic formulation requires that $d$ be everywhere positive, and
the uncertainties will be truncated to only allow positive values and
incorrect in detail.  However, at high signal to noise they should be
approximately correct. {\color{red} As I understand it, additional
`bad lines' in the mean model could be accounted for using narrow
quadratics in the covariance matrix, though their interpretation would
be hellish.}

@@

@@

We first infer the parameters of interest $\theta$ for a calibrated
spectrum of one of the clusters, \excluster, using the full YY-band
photometry. In Figure \ref{fig:inferred_spectrum} we show the spectra
that result from draws of the parameters $[\theta,\alpha]$ from the
posterior distribution, as well as several of the individual
components of the model constructed from each draw. Figure
\ref{fig:inferred_spectrum} also shows the calibrated spectrum on
which the priors were conditioned, and residuals between the posterior
models and the observed data.  The agreement is quite good, though
some lines are not accurately predicted.  For example, the observed
Mg$b$ absorption appears stronger than the modeled
absorption. {\color{blue} Should I show blowups of selected
regions/lines?}

@@

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{figures/pfig_b192-g242_225_cal.png}
\caption{ Model SEDs.  Top: The model photometric SED resulting from
draws from the posterior parameter space conditioned ({\it cyan}) on
the observed photometry ({\it magenta}) and spectra.  Bottom:
uncertainty normalized residuals of the models from the observations.
\label{fig:inferred_sed}}
\end{figure}

\begin{figure}[h!]
\includegraphics[width=0.5\textwidth]{figures/ptri_b192-g242_225_cal.png}
\caption{ Marginalized and joint PDFs of subset of the physical
parameters $\theta$ inferred from the spectophotometrically calibrated
spectrum of \excluster. The blue lines indicate the parameter values
that resulted from the powell minimization.
\label{fig:inferred_params}}
\end{figure}

@@

For the calibration model, we obtain a vector $f(\alpha)$
that is nearly linear in wavelength {\color{red} WHY IS IT NOT A
CONSTANT?  Need longer burn-in?  Is this the same issue as the SED
residuals?}.  The length scale of the Gaussian process is $\sim
60$\AA.  It is possible to reconstruct the inferred total
multiplicative calibration vector for each draw of the model
parameters.  Rearranging equation \ref{eq:spectroscopicLF} yields
\begin{eqnarray} 
\frac{d}{\mu(\theta)} & = & f(\alpha) +
\frac{\tilde{\Delta}}{\mu(\theta)}
\quad ,
\end{eqnarray}
for the total effective multiplicative calibration, where
$\tilde{\Delta}$ is now the Gaussian process mean prediction for the
residual between the data and the model. The inverse of this ratio is
shown in the left panel of Figure \ref{fig:inferred_calib} as our
inferred total multiplicative calibration function; if the
spectrophotometric calibration were perfect and we had correctly
inferred it, then we would expect this vector to be constant as a
function of wavelength.

\begin{figure*}[h!]
\includegraphics[width=\textwidth]{figures/cfig_b192-g242_225.png}
\caption{Left: Samples of the effective multiplicative calibration
vector for the spectrophotometrically calibrated spectrum of
\excluster~ ({\it cyan}) compared to that obtained from ancillary
photometric data ({\it magenta}).  
%Right: The same, but for the uncalibrated data.
 \label{fig:inferred_calib}}
\end{figure*}

\subsection{Tests}

BDJ: Next, we infer the parameters of the spectrum of the same cluster
\emph{without} spectrophotometric calibration.  How many/which figures
to show?

BDJ: Then, we infer the parameters of the calibrated spectrum without
including the noise model!  What changed, what did we get wrong
without the noise model?

BDJ: We should also test on fake data

BDJ: Then, we show select marginalized or joint PDFs for each cluster
calibrated/uncalibrated, to assess consistency.  Or we just show
comparisons of confidence intervals.

\begin{figure}[h!]
\includegraphics[width =0.5\textwidth]{figures/B192_020_cratio.pdf}
\caption{Money plot.  Show the difference between the parameters
  inferred from the calibrated and uncalibrated spectra, for each
  cluster. Tjis is a placeholder, since it only shows B192 for fixed
  GP parameters. \label{fig:boom}}
\end{figure}

@@

We have demonstrated, using a sample of stellar clusters in M31, that
we can combine photometric and spectroscopic data in a principled way
to infer physical properties.  We do not resort to an arbitrary
down-weighting or up-weighting of certain data. Instead, we explicitly
model our uncertain knowledge of the spectral calibration and infer
this calibration jointly with the physical properties. For a number of
applications this combination of spectroscopy and photometry reduces
the need for spectrophotometric calibration entirely, such that the less
expensive photometry is sufficient.  Additional calibration
information, if it exists, can be incorporated into informative priors
on the calibration parameters of the model, to inform the construction
of the calibration model, or to approximately remove gross features
that are difficult to model parameterically or as a sum of Gaussians.

@@

% Fundamentally, calibration involves observing a `known' or `truth'
% with an instrument, and using the observation to infer the effects of
% the instrument.  However, we do not have knowledge of a single truth
% for any object, only relative probabilities for what might be true.
% In this sense, one should never use a single point estimate of the
% calibration, the full distribution of the possible calibrations should
% be inferred from our probabilistic knowledge of the calibration object
% and forward propogated \citep[e.g.][]{lee11}. Once this is done, the only the
% only distinction between 'calibration' objects and targets of interest
% is the complexity of the model for the object and the level of
% uncertainty in the calibration object spectrum.  If one is interested
% in inferring parameters of interest of spectral models of
% astrophysical objects, then there is no reason not to use those
% objects themselves as calibration objects.

>>>>

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a trivial modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N \left[\frac{[d_n - \mu_n(\theta)]^2}{[\sigma_n^2 + s^2]} + \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the
log-likelihood function we are going to use for the photometric
measurements ({\color{blue} BDJ: IS THIS TRUE?} {\color{red} Jitter
will be added now with correct sign as I test increases in model
complexity }) in what follows.  
In this formulation, if $\theta$ are the parameters of interest, the
jitter variance $s^2$ is a nuisance noise-model parameter.  The
Gaussian Process will be a generalization of this modification.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& \ln d - \ln \mu(\theta) - f(\alpha) 
\\
C_{nn'} &\equiv& [\hat{\sigma}_n^2 + s^2]\,\delta_{nn'} +
K_\alpha(\lambda_n - \lambda_{n'})
\\
\hat{\sigma_n} & = &\sigma_n/d
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
the ``$\ast$'' operator is the pixel-wise multiplication operator,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function,
$\hat{\sigma}_n$ are the fractional uncertainties of the data,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian process is going to
pick up smaller-scale residuals away from \emph{that}.
The residual column vector is defined in terms of the logarithmic
difference between the data and the mean model $\mu$ since we are
taking the calibration vector to be multiplicative with respect to
the mean model $\mu$.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In this formulation the the mean model $\mu$ will be effectively
multiplied by $e ^{f(\alpha)}$ ({\color{red} could we just add this to
the covariance matrix with a single long quadratic ?}).  This
logarithmic formulation requires that $d$ be everywhere positive, and
the uncertainties will be truncated to only allow positive values and
are therefore incorrect in detail.  However, at high signal to noise
the results should be unaffected. {\color{red} As I understand it,
additional `bad lines' in the mean model could be accounted for using
narrow quadratics in the covariance matrix, though their
interpretation would be hellish. The best way is probably to explictly
model them in the mean model}

@@

\subsection{List of Tests}

\begin{enumerate}

\item We should test on fake data, with known squiggly
  calibration applied, to demonstrate that we can infer the
  calibration vector and the physical parameters with no bias.

\item We infer the parameters of the spectrum of a real cluster that
  has been photometrically calibrated.

\item Next, we infer the parameters of the spectrum of the same
cluster \emph{without} spectrophotometric calibration, to see if we
infer physical parameters consistent with the calibrated case.  How
many/which figures to show?

\item  Then, we show select marginalized or joint PDFs for many
clusters calibrated/uncalibrated, to assess consistency.  Or we just
show comparisons of confidence intervals for the parameters of a
number of clusters.

\item We infer the parameters of the calibrated spectrum without
including the noise model, or presumably the photometry (since
reconciling photometry and spectroscopy is done by our noise model).
What changed, what did we get wrong without the noise model?
{\color{blue} Should we keep the $f(\alpha)$ and the photometry?  This
just has the potential to show shortcomings in the spectrophotometric
calibration calibration}

\item Finally, we change the number of photometric bands and see what
happens to the inferred parameters, using the full model.

\end{enumerate}

@@

We first infer the parameters of interest $\theta$ for a calibrated
spectrum of one of the clusters, \excluster, using the full YY-band
photometry. In Figure \ref{fig:inferred_spectrum} we show the spectra
that result from draws of the parameters $[\theta,\alpha]$ from the
posterior distribution, as well as several of the individual
components of the model constructed from each draw. Figure
\ref{fig:inferred_spectrum} also shows the calibrated spectrum on
which the priors were conditioned, and residuals between the posterior
models and the observed data.  The agreement is quite good, though
some lines are not accurately predicted.  For example, the observed
Mg$b$ absorption appears stronger than the modeled
absorption.

@@

\begin{figure}[h!]
%\includegraphics[width=0.5\textwidth]{figures/pfig_b192-g242_225_cal.png}
\caption{ Model SEDs.  Top: The model photometric SED resulting from
draws from the posterior parameter space conditioned ({\it cyan}) on
the observed photometry ({\it magenta}) and spectra.  Bottom:
uncertainty normalized residuals of the models from the observations.
\label{fig:inferred_sed}}
\end{figure}

\begin{figure}[h!]
%\includegraphics[width=0.5\textwidth]{figures/ptri_b192-g242_225_cal.png}
\caption{ Marginalized and joint PDFs of subset of the physical
parameters $\theta$ inferred from the spectophotometrically calibrated
spectrum of \excluster. The blue lines indicate the parameter values
that resulted from the powell minimization.
\label{fig:inferred_params}}
\end{figure}

@@

It is possible to reconstruct the inferred total multiplicative
calibration vector for each draw of the model parameters.  Rearranging
equation \ref{eq:spectroscopicLF} yields
\begin{eqnarray} 
\frac{d}{\mu(\theta)} & = & exp(f(\alpha) + \tilde{\Delta})
\quad ,
\end{eqnarray}
for the total effective multiplicative calibration, where
$\tilde{\Delta}$ is now the Gaussian process mean prediction for the
logarithmic residual between the data and the model. The inverse of
this ratio is shown in the left panel of Figure
\ref{fig:inferred_calib} as our inferred total multiplicative
calibration function; if the spectrophotometric calibration were
perfect and we had correctly inferred it, then we would expect this
vector to be constant as a function of wavelength.

\begin{figure*}[h!]
%\includegraphics[width=\textwidth]{figures/cfig_b192-g242_225.png}
\caption{Left: Samples of the effective multiplicative calibration
vector for the spectrophotometrically calibrated spectrum of
\excluster~ ({\it cyan}) compared to that obtained from ancillary
photometric data ({\it magenta}).  
%Right: The same, but for the uncalibrated data.
 \label{fig:inferred_calib}}
\end{figure*}

\begin{figure}[h!]
%\includegraphics[width =0.5\textwidth]{figures/B192_020_cratio.pdf}
\caption{Money plot.  Show the difference between the parameters
  inferred from the calibrated and uncalibrated spectra, for each
  cluster. Tjis is a placeholder, since it only shows B192 for fixed
  GP parameters. \label{fig:boom}}
\end{figure}

@@

We have demonstrated, using a sample of stellar clusters in M31, that
we can combine photometric and spectroscopic data in a principled way
to infer physical properties.  We do not resort to an arbitrary
down-weighting or up-weighting of certain data. Instead, we explicitly
model our uncertain knowledge of the spectral calibration and infer
this calibration jointly with the physical properties. For a number of
applications this combination of spectroscopy and photometry reduces
the need for spectrophotometric calibration entirely, such that the less
expensive photometric calibration is sufficient.  Additional calibration
information, if it exists, can be incorporated into informative priors
on the calibration parameters of the model, inform the construction
of the calibration model, or approximately remove gross features
that are difficult to model parameterically or as a sum of Gaussians.

{\color{blue} BDJ words about the gains that might be made by
including photometry over a large range in wavelngth with spectroscopy
over relatively small ranges.}

@@

{\color{blue} BDJ - This paragraph is probably too calibration
focused. We want to focus more on spec + phot combinations.  At least,
this paragraph should not be in the discussion section.}
Fundamentally, calibration involves observing a `known' or `truth'
with an instrument, and using the observation to infer the effects of
the instrument.  However, we do not have knowledge of a single truth
for any object, only relative probabilities for what might be true.
In this sense, one should never use a single point estimate of the
calibration, the full distribution of the possible calibrations should
be inferred from our probabilistic knowledge of the calibration object
and forward propogated \citep[e.g.][]{lee11}. Once this is done, the only the
only distinction between 'calibration' objects and targets of interest
is the complexity of the model for the object and the level of
uncertainty in the calibration object spectrum.  If one is interested
in inferring parameters of interest of spectral models of
astrophysical objects, then there is no reason not to use those
objects themselves as calibration objects.
